\appendix
\section{Appendix: Preliminary Performance Benchmarks}
\label{app:benchmarks}

The following tables summarize preliminary performance benchmarks for key DeepCausality operations, executed on an Apple MacBook Pro with an M3 Max processor. These benchmarks were run using the standard Rust `criterion` library (`cargo bench`). The benchmark code is available on Github\footnote{https://github.com/deepcausality-rs/deep\_causality/tree/main/deep\_causality/benches}. It is important to note that these are synthetic benchmarks, primarily utilizing linear or simple multi-layer graph structures, and do not involve complex context interactions or GPU acceleration. They provide an initial indication of the framework's CPU-based performance characteristics at different scales.

The scales are defined as follows:
\begin{itemize}
    \item Small: 10 Causaloids (for collections/maps/graphs)
    \item Medium: 1,000 Causaloids
    \item Large: 10,000 Causaloids
\end{itemize}

Table~\ref{tab:bench_collections} presents results for reasoning over collections (Vector/Array) and Maps (HashMap/BTreeMap) containing Causaloids. Table~\ref{tab:bench_linear_graphs} details performance for various reasoning tasks on linear `CausaloidGraph` structures. Table~\ref{tab:bench_multilayer_graphs} shows results for reasoning on small multi-layer graphs (benchmarks for medium/large multi-layer graphs were not included in the provided results). All times reported are the typical execution times (median/mean) derived from multiple samples.

% Table for Collection and Map Reasoning
\begin{table}[htbp]
    \centering
    \caption{Benchmark Results: Reasoning over Causaloid Collections and Maps.}
    \label{tab:bench_collections}
    \begin{tabular}{llr}
        \toprule
        Operation Description                 & Scale (Size) & Typical Time \\
        \midrule
        Collection (`Vec`/`Array`) Reasoning  & Small (10)   & \SI{101.01}{ns} \\
                                              & Medium (1k)  & \SI{3.9123}{\micro s} \\
                                              & Large (10k)  & \SI{43.979}{\micro s} \\
        \midrule
        Map (`HashMap`/`BTreeMap`) Reasoning & Small (10)   & \SI{50.365}{ns} \\
                                              & Medium (1k)  & \SI{4.6652}{\micro s} \\
                                              & Large (10k)  & \SI{50.797}{\micro s} \\
        \bottomrule
    \end{tabular}
\end{table}

% Table for Linear Graph Reasoning
\begin{table}[htbp]
    \centering
    \caption{Benchmark Results: Reasoning over Linear Causaloid Graphs.}
    \label{tab:bench_linear_graphs}
    \begin{tabular}{llr}
        \toprule
        Reasoning Task on Graph           & Scale (Nodes) & Typical Time \\
        \midrule
        Reason over All Causes            & Small (10)    & \SI{2.7601}{\micro s} \\
                                          & Medium (1k)   & \SI{509.94}{\micro s} \\
                                          & Large (10k)   & \SI{70.221}{ms} \\
        \midrule
        Reason over Subgraph from Cause   & Small (10)    & \SI{1.5070}{\micro s} \\
                                          & Medium (1k)   & \SI{245.25}{\micro s} \\
                                          & Large (10k)   & \SI{34.933}{ms} \\
        \midrule
        Reason Shortest Path Between Causes & Small (10)    & \SI{1.6900}{\micro s} \\
                                          & Medium (1k)   & \SI{286.08}{\micro s} \\
                                          & Large (10k)   & \SI{35.424}{ms} \\
        \midrule
        Reason over Single Cause          & Small (10)    & \SI{10.349}{ns} \\
                                          & Medium (1k)   & \SI{9.9537}{ns} \\
                                          & Large (10k)   & \SI{10.010}{ns} \\
        \bottomrule
    \end{tabular}
\end{table}

\newpage

% Table for Multi-Layer Graph Reasoning (Small Scale Only)
\begin{table}[htbp]
    \centering
    \caption{Benchmark Results: Reasoning over Small Multi-Layer Causaloid Graphs.}
    \label{tab:bench_multilayer_graphs}
    \begin{tabular}{llr}
        \toprule
        Reasoning Task on Graph           & Scale (Nodes) & Typical Time \\
        \midrule
        Reason over All Causes            & Small (10)    & \SI{1.2483}{\micro s} \\
        Reason over Subgraph from Cause   & Small (10)    & \SI{489.42}{ns} \\
        Reason Shortest Path Between Causes & Small (10)    & \SI{427.45}{ns} \\
        Reason over Single Cause          & Small (10)    & \SI{10.252}{ns} \\
        \bottomrule
    \end{tabular}
\end{table}

These preliminary results demonstrate efficient CPU-bound performance for reasoning tasks on the tested structures and scales, particularly highlighting the near-constant O(1) time for single cause lookups, irrespective of graph size. The scaling for full graph traversals appears roughly linear or slightly super-linear for the linear graphs tested up to 10,000 nodes. However, as noted in the Discussion, further benchmarking on more complex, densely connected hypergraphs and under various context interaction scenarios is necessary to fully characterize scalability limits. The raw benchmark output also indicated the presence of outliers in some measurements, suggesting potential sources of variability (e.g., cache effects, OS scheduling) that warrant deeper investigation in future performance analyses.
