\documentclass{article}

\usepackage{../styles/PRIMEarxiv} % imports style
\input{../shared/package} % imports packages 

% imports code formatting used across the document 
\input{../styles/rust-code-style.tex}

%Formating code listings
% https://nasa.github.io/nasa-latex-docs/html/examples/listing.html
\usepackage{listings} 
\lstset
{ %Formatting for code in appendix
    basicstyle=\footnotesize,
    numbers=left,
    stepnumber=1,
    showstringspaces=false,
    tabsize=1,
    breaklines=true,
    breakatwhitespace=false,
}

%Header
\pagestyle{fancy}
\thispagestyle{empty}
\rhead{ \textit{ }} 

% Update your Headers here
\fancyhead[LO]{The Effect Propagation Process (EPP): An Axiomatic Foundation for Dynamic Causality}
  
%% Title
\title{The Effect Propagation Process (EPP):\newline An Axiomatic Foundation for Dynamic Causality}

%% Author
\author{
  Marvin Hansen \\
  \texttt{marvin.hansen@gmail.com} \\
   Date: \today
}


\begin{document}
\maketitle

\begin{abstract}
Contemporary frameworks for computational causality provide a formal basis for inference within systems governed by static causal structures. These models, however, are predicated on assumptions of a fixed background spacetime and linear temporal progression, which become fundamental limitations when addressing complex dynamic systems where the causal laws themselves can evolve.

This monograph introduces the Effect Propagation Process (EPP), a formal meta-calculus for dynamic causality. It provides a single axiomatic foundation, a formal language, and a set of computable primitives for constructing, composing, and verifying domain-specific, causal theories. The EPP's single axiomatic foundation that causality is a spacetime-agnostic functional dependency necessitates a trifecta of core architectural components. The Context is an explicit and dynamic hypergraph that models the operational environment, supporting non-Euclidean and non-linear temporal structures. The Causaloid is a polymorphic, composable unit of causal mechanism that can encapsulate any specific object-level calculus. The Effect Ethos is a  deontic guardrail that uses a defeasible calculus to ensure a system's actions remain verifiably aligned with its safety and mission objectives.

The EPP is fully implemented in the open-source DeepCausality peoject, demonstrating the work's practical application. Together, the EPP and its implementation in DeepCausality provide a foundation for the design, simulation, and implementation of a new generation of robust, context-aware, and verifiably aligned dynamic causal systems.

\end{abstract}

\newpage

%% Table of Contents
\tableofcontents

\newpage

%% Introduction
\input{introduction.tex}

%% History
\input{../epp_chapters/history.tex}

%% Related Work
\input{../epp_chapters/related_work.tex}

%% Effect Propagation Process
\input{../epp_chapters/causality_as_epp.tex}

%% Metaphysics
\input{../epp_chapters/metaphysics.tex}

%% Epistemology
\input{../epp_chapters/epistemology.tex}

%% Teleology
\input{../epp_chapters/teleology.tex}

%% Ontology
\input{../epp_chapters/ontology.tex}

%% Formalization  
\input{../epp_chapters/formalization.tex}

%% Validity 
%%\input{../epp_chapters/validity.tex}

%% Implementation 
\input{../epp_chapters/implementation.tex}
\input{../epp_chapters/implementation_ultragraph.tex}
\input{../epp_chapters/implementation_deep_causality.tex}
\input{../epp_chapters/implementation_benchmarks.tex}
\input{../epp_chapters/implementation_validation.tex}

%% Future Work
\input{../epp_chapters/future_work.tex}

%% Epilogue
\input{../epp_chapters/epilogue.tex}

%% Appendix A
\input{../epp_appendices/appendix_a.tex}

%% Glossary
\input{../epp_chapters/glossary.tex}

%% Bibliography
\newpage
\bibliographystyle{unsrt}  
\addcontentsline{toc}{section}{References} % Add it to the Table of Contents
\bibliography{../epp_bib/references}  

\end{document}
